### 2019-11-26 ic-journal/multivariate.tex
text

In this section,  we discuss the formalisation of the \multivariate
cases of Milner's ``unique solution of equations (for $\sim$)''
(Theorem~\ref{t:Mil89s1}) and our ``unique solution of rooted
contractions'' (Theorem~\ref{t:rcontraBisimulationU}).
\hl{This formalisation only supports finitely many
  equations/contractions and equation variables.}\footnote{\hl{The original theorems
hold for even (uncountably) infinite equations and equation variables.}}
We chose these two theorems because they both depend on the same
concept of ``weakly guarded contexts'', and Lemma~\ref{lem:milner313}.
The work is an extension of the \univariate case~\cite{EPTCS276.10}
 in the sense that neither the CCS datatype nor
the SOS rules need to be changed.

@@

In the  literature the variables  $X_i$ of
a system of equations $\{X_i = E_i\}_{i\in I}$ are usually considered as
\emph{equation variables}
\emph{outside} of the CCS syntax; then an equation body  $E_i$  is an \emph{open} expression
built with CCS operators plus the equation variables.
For a formal point of view, this means we either need to define a
whole new datatype (with extra equation variables) in which each CCS
opeartor must be duplicated, or we have to  add equation variables as a
new primitive (e.g.~\texttt{Var X}) to the existing CCS type (besides
the agent variables such as  \texttt{var X}).
 Then the substitution of equation
variables in $E_i\{\til P/\til X\}$ would be the \emph{syntactic substitution} of
each occurrence of $X_i$ in $E_i$ with the corresponding $P_i$.
In either cases the \emph{disjointness} between equation variables and
agent variables is syntactically
guaranteed. Both solutions are rather combersome, and require a non-trivial modification
of the formalisation of the \univariate case.

In this project, we have followed Milner~\cite{Mil89}, using the
same alphabet for both agent variables and equation variables.
This has allowed a large reuse of the \univariate formalisation.
For instance, the variable
substitution function used in SOS rules \texttt{REC} can be also
used for substituting eqaution variables.
However, a lot of care is needed in the proofs of many
fundamental lemmas, mostly due to the variable capture issues between
equation variables and agent variables.

@@

Recall \hl{from} the beginning of Section~\ref{s:eq} \hl{that},
CCS syntax leaves the syntactic possibility that an
agent variable $A$ occurs outside the recursion operator of the same
variable. In Milner's original book chapter~\cite{milner1990operational}, the
variables which occur in Recursion (such as $X$ in $\recu X E$) are
called \emph{bound} variables, while those which occur unbound are called
\emph{free} variables.

We denote the set of all bound variables of a
given CCS term $E$ (we recall that these are variables that are bound in a recursion
subexpression of $E$) as $\bv{E}$ (or \HOLinline{\HOLConst{BV}\\\;\HOLFreeVar{E}} in HOL), and the set of  free
variables as $\fv{E}$ (or \HOLinline{\HOLConst{FV}\\\;\HOLFreeVar{E}} in HOL). Both \HOLinline{\HOLConst{BV}} and \HOLinline{\HOLConst{FV}} have the
type ``\HOLinline{\ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS} \HOLTokenTransEnd \ensuremath{\alpha} \HOLTokenTransEnd \HOLTyOp{bool}}'', i.e. functions taking CCS terms returning sets of
variables (of type \HOLinline{\ensuremath{\alpha}}).
% Milner didn't give their definitions explicitly. However, one can
% imagine that, for most CCS operators they just pass into
% subterms without collecting anything (and $\nil$ gives $\emptyset$).
For their definition the interesting cases are those of 
recursion and agent variables, shown below  (here \texttt{DELETE}
and \texttt{INSERT} are set-theoretic operators of HOL's
\texttt{pred_set} theory):
\begin{center}
\begin{tabular}{|l|l|}
\hline
\HOLConst{FV} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLTokenDefEquality{} \HOLTokenLeftbrace{}\HOLFreeVar{X}\HOLTokenRightbrace{} & \HOLConst{FV} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{p}\ensuremath{)} \HOLTokenDefEquality{} \HOLConst{FV} \HOLFreeVar{p} \HOLConst{DELETE} \HOLFreeVar{X} \\
\HOLConst{BV} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLTokenDefEquality{} \HOLSymConst{\HOLTokenEmpty{}} & \HOLConst{BV} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{p}\ensuremath{)} \HOLTokenDefEquality{} \HOLFreeVar{X} \HOLConst{INSERT} \HOLConst{BV} \HOLFreeVar{p} \\
\hline
\end{tabular}
\end{center}
Furthermore, $E$ is a process, written \HOLinline{\HOLConst{IS_PROC} \HOLFreeVar{E}}, if
it does not contain any free variable, i.e. $\fv{E} = \emptyset$:
\begin{alltt}
   \HOLConst{IS_PROC} \HOLFreeVar{E} \HOLTokenDefEquality{} \HOLConst{FV} \HOLFreeVar{E} \HOLSymConst{\ensuremath{=}} \HOLSymConst{\HOLTokenEmpty{}}\hfill{[IS_PROC_def]}
\end{alltt}
And a list of CCS processes can be asserted by \HOLinline{\HOLConst{ALL_PROC}} defined upon \HOLinline{\HOLConst{IS_PROC}}:
\begin{alltt}
   \HOLConst{ALL_PROC} \HOLFreeVar{Es} \HOLTokenDefEquality{} \HOLConst{EVERY} \HOLConst{IS_PROC} \HOLFreeVar{Es}\hfill{[ALL_PROC_def]}
\end{alltt}

Note that, in CCS expression, the set of free and bound
variables need not be disjoint, e.g.\ in  $X + \recu X E$.
More importantly, when going
from a CCS expression to its sub-expressions, the set of free
variables may increase, while the set of bound variables either
remains the same or decreases. For instance, $\fv{\recu X (\mu.X)} = \emptyset$, while
$\fv{\mu.X} = \{X\}$. This property of $\fv{\cdot}$ brings some
difficulities when doing proofs about CCS transitions. As an evidence,
we prove the following fundamental result about $\fv{\cdot}$\hl{\mbox{~\citep[p.~1209]{milner1990operational}}}:

\begin{proposition}
\label{prop:transFV}
The derivatives of a process are themselves processes, i.e.
if $E \arr{\mu} E'$ and $\fv{E} = \emptyset$, then $\fv{E'} =
\emptyset$. Or, formally:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{E} \HOLBoundVar{u} \ensuremath{\HOLBoundVar{E}\sp{\prime}}. \HOLBoundVar{E} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{IS_PROC} \HOLBoundVar{E} \HOLSymConst{\HOLTokenImp{}} \HOLConst{IS_PROC} \ensuremath{\HOLBoundVar{E}\sp{\prime}}\hfill{[TRANS_PROC]}
\end{alltt}
\end{proposition}

\begin{proof}
We \hl{actually proved} a stronger result \hl{saying} the set of free
variables \hl{in any CCS process} decreases through its derivatives:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{E} \HOLBoundVar{u} \ensuremath{\HOLBoundVar{E}\sp{\prime}}. \HOLBoundVar{E} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{FV} \ensuremath{\HOLBoundVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{FV} \HOLBoundVar{E}\hfill{[TRANS_FV]}
\end{alltt}

\hl{Milner said it can be proven by ``an easy action induction''.}
\hl{It is actually not that easy, and we take this chance to
  illustrate how to apply ``action induction'' (in HOL), which} is actually a higher-order application
of the following \emph{induction principle} generated together with SOS
rules (which essentially says that \HOLinline{\HOLConst{TRANS}} is the smallest relation
satisfying SOS rules):

@@

would have completed.
This is the essence of action induction.
Now the original goal can be reduced to several conjunct

@@

the following goal left: (the term above the dash line is the goal,
those below the line are assumptions)

@@

Notice that, \HOLinline{\HOLConst{FV}\\\;\HOLFreeVar{E}\\\;\HOLConst{DELETE}\\\;\HOLFreeVar{X}\\\;\HOLSymConst{\ensuremath{=}}\\\;\HOLConst{FV}\\\;\ensuremath{(}\HOLConst{rec}\\\;\HOLFreeVar{X}\\\;\HOLFreeVar{E}\ensuremath{)}}, so

@@

would have completed. Thus it remains to show that

@@

Proposition~\ref{prop:transFV} \hl{is essential in the proofs of the
  multivariable versions of all unique-solution
  theorems.
In comparison with} the above proof, the proof of the similar
result for bound variables is \hl{much} easier:
\begin{proposition}
\label{prop:transBV}
if $E \arr{\mu} E'$ then $\bv{E'} \subseteq \bv{E}$, or formally:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{E} \HOLBoundVar{u} \ensuremath{\HOLBoundVar{E}\sp{\prime}}. \HOLBoundVar{E} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{BV} \ensuremath{\HOLBoundVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{BV} \HOLBoundVar{E}\hfill{[TRANS_BV]}
\end{alltt}
\end{proposition}
\begin{proof}
Following the proof of Prop.~\ref{prop:transFV}, we do action
induction on $E \arr{\mu} E'$ and simplify all subgoals by set-theoretic
facts. The only subgoal left is:
\begin{alltt}
        \HOLinline{\HOLConst{BV} \ensuremath{\HOLFreeVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLFreeVar{X} \HOLConst{INSERT} \HOLConst{BV} \HOLFreeVar{E}}
   ------------------------------------
    0.  \HOLinline{\HOLConst{BV} \ensuremath{\HOLFreeVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{BV} \ensuremath{(}\ensuremath{[}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)}}
\end{alltt}
Note the difference with the proof of Prop.~\ref{prop:transFV}: now
we have \texttt{INSERT} in place of \texttt{DELETE}. Then, we directly
use the assumption and the transitivity of $\subseteq$ and try to
prove the following goal instead:
\begin{alltt}
        \HOLinline{\HOLConst{BV} \ensuremath{(}\ensuremath{[}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)} \HOLSymConst{\HOLTokenSubset{}} \HOLFreeVar{X} \HOLConst{INSERT} \HOLConst{BV} \HOLFreeVar{E}}
   ------------------------------------
    0.  \HOLinline{\HOLConst{BV} \ensuremath{\HOLFreeVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{BV} \ensuremath{(}\ensuremath{[}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)}}
\end{alltt}
This is an easy goal, due to the following lemma which is
 proved by induction on $E$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{X} \HOLBoundVar{E} \ensuremath{\HOLBoundVar{E}\sp{\prime}}. \HOLConst{BV} \ensuremath{(}\ensuremath{[}\ensuremath{\HOLBoundVar{E}\sp{\prime}}\ensuremath{/}\HOLBoundVar{X}\ensuremath{]} \HOLBoundVar{E}\ensuremath{)} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{BV} \HOLBoundVar{E} \HOLSymConst{\HOLTokenUnion{}} \HOLConst{BV} \ensuremath{\HOLBoundVar{E}\sp{\prime}}\hfill{[BV_SUBSET]}
\end{alltt}
Taking $E ' = \recu X E$ in the above lemma, we have
    \HOLinline{\HOLConst{BV}\\\;\ensuremath{(}\ensuremath{[}\HOLConst{rec}\\\;\HOLFreeVar{X}\\\;\HOLFreeVar{E}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]}\\\;\HOLFreeVar{E}\ensuremath{)}\\\;\HOLSymConst{\HOLTokenSubset{}}\\\;\HOLConst{BV}\\\;\HOLFreeVar{E}\\\;\HOLSymConst{\HOLTokenUnion{}}\\\;\HOLConst{BV}\\\;\ensuremath{(}\HOLConst{rec}\\\;\HOLFreeVar{X}\\\;\HOLFreeVar{E}\ensuremath{)}},
    the right side equals to \HOLinline{\HOLConst{BV}\\\;\HOLFreeVar{E}\\\;\HOLSymConst{\HOLTokenUnion{}}\\\;\ensuremath{(}\HOLFreeVar{X}\\\;\HOLConst{INSERT}\\\;\HOLConst{BV}\\\;\HOLFreeVar{E}\ensuremath{)}} $=$
    \HOLinline{\HOLFreeVar{X}\\\;\HOLConst{INSERT}\\\;\HOLConst{BV}\\\;\HOLFreeVar{E}}. Thus
the    proof can 
 be completed.
\end{proof}

@@

There are two ways to represent  the \multivariate
substitution of $E\{\til P/\til X\}$ which substitutes each \hl{possible
occurrence} of free variables $X_i$ in $E$ with the corresponding
$P_i$: (1) iterately applying the existing \univariate
\HOLinline{\HOLConst{CCS_Subst}}; (2) defining a new \multivariate substitution
function.
% In any case, we (reasonably) assume that all $X_i$ are
%distinct with each other. 
Note that  there is the
possibility that $\til P$   syntactically  again contains variables from $\til X$, thus
different orders of iterated substitutions may lead to different results.
% In fact, there are all kinds of issues with
% substitutions and applying them to term-like structures. 
To obtain the
maximal flexibility with a reduced effort, we defined
a \multivariate version of \HOLinline{\HOLConst{CCS_Subst}}
called \HOLinline{\HOLConst{CCS_SUBST}} based HOL's \texttt{finite\_map} theory~\cite{holdesc}.

@@

The function \HOLinline{\HOLConst{CCS_SUBST}} takes a finite
map \HOLinline{\HOLFreeVar{fm}} of type \HOLinline{\ensuremath{\alpha} \HOLTokenMapto{} \ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS}} and a CCS expression,
and then returns another CCS expression in which
all occurrences of variables in the \HOLinline{\HOLFreeVar{fm}}'s domain are substituted with
the corresponding value in \HOLinline{\HOLFreeVar{fm}}. Such a finite map can be built from
the list of variables \HOLinline{\HOLFreeVar{Xs}} and the corresponding targets
\HOLinline{\HOLFreeVar{Ps}} (of the same length) by a helper function \HOLinline{\HOLConst{fromList}} (whose details are
not interesting and omitted here), while
\HOLinline{\HOLConst{CCS_SUBST}\\\;\ensuremath{(}\HOLConst{fromList}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Ps}\ensuremath{)}\\\;\HOLFreeVar{E}} is
abbreviated as \HOLinline{\ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]}\\\;\HOLFreeVar{E}}.
The substitution process is
variable-parallel thus order-independent. For most CCS operators it
just recursively calls itself on subterms. The only
interesting cases are at agent variables and recursion:
\begin{alltt}
   \HOLConst{CCS_SUBST} \HOLFreeVar{fm} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLTokenDefEquality{} \HOLKeyword{if} \HOLFreeVar{X} \HOLSymConst{\HOLTokenIn{}} \HOLConst{FDOM} \HOLFreeVar{fm} \HOLKeyword{then} \HOLFreeVar{fm} \HOLConst{'} \HOLFreeVar{X} \HOLKeyword{else} \HOLConst{var} \HOLFreeVar{X}\hfill{[CCS_SUBST_var]}

>>>>

In this section,  we discuss the formalisation of the \multivariate
cases of Milner's ``unique solution of equations (for $\sim$)''
(Theorem~\ref{t:Mil89s1}) and our ``unique solution of rooted
contractions'' (Theorem~\ref{t:rcontraBisimulationU}).
\hl{This formalisation  supports finitely many
  equations/contractions and equation variables.}\footnote{\hl{The original theorems
hold for even  infinitely-many equations and equation variables.}}
We chose these two theorems because of their relevance and 
because they well illustrate the effort needed for the \multivariate case. 
% Both theorems depend on the same
% concept of ``weakly guarded contexts'', and Lemma~\ref{lem:milner313}.
The work is an extension of the \univariate case %~\cite{EPTCS276.10}
 in the sense that neither the CCS datatype nor
the SOS rules need to be changed.

@@

In the  literature the variables  $X_i$ of
a system of equations $\{X_i = E_i\}_{i\in I}$ are usually considered as
equation variables
\emph{outside} of the CCS syntax; then an equation body  $E_i$  is an \emph{open} expression
built with CCS operators plus the equation variables.
For a formal point of view, this means we either need to define a
whole new datatype (with extra equation variables) in which each CCS
operator must be duplicated, or we have to  add equation variables as a
new primitive (e.g.~\texttt{Var X}) to the existing CCS type (besides
the agent variables such as  \texttt{var X}).
 Then the substitution of equation
variables in $E_i\{\til P/\til X\}$ would be the \emph{syntactic substitution} of
each occurrence of $X_i$ in $E_i$ with the corresponding $P_i$.
In either cases the \emph{disjointness} between equation variables and
agent variables is syntactically
guaranteed. Both solutions are rather combersome, and require a non-trivial modification
of the formalisation of the \univariate case.

In this project, we have followed Milner~\cite{Mil89}, using the
same alphabet for both agent variables and equation variables.
This has allowed a large reuse of the \univariate formalisation.
For instance, the variable
substitution function used in SOS rules \texttt{REC} can be also
used for substituting equation variables.
However, a lot of care is needed in the proofs of many
fundamental lemmas, mostly due to the variable capture issues between
equation variables and agent variables.

@@

As mentioned at
 the beginning of Section~\ref{s:eq},  
 an
agent variable may  occur outside the recursion in which it is bound;
i.e., a variable can appear both  \emph{bound} and \emph{free} in a term. 
% In Milner's original book chapter~\cite{milner1990operational}, the
% variables which occur in Recursion (such as $X$ in $\recu X E$) are
% called \emph{bound} variables, while those which occur unbound are called
% \emph{free} variables.

We denote the set of bound variables of a
given CCS term $E$ (the variables that are bound in a recursion
subexpression of $E$) as $\bv{E}$ (or \HOLinline{\HOLConst{BV}\\\;\HOLFreeVar{E}} in HOL), and the set of  free
variables as $\fv{E}$ (or \HOLinline{\HOLConst{FV}\\\;\HOLFreeVar{E}} in HOL). Both \HOLinline{\HOLConst{BV}} and \HOLinline{\HOLConst{FV}} have the
type ``\HOLinline{\ensuremath{(}\ensuremath{\alpha},
  \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS} \HOLTokenTransEnd
  \ensuremath{\alpha} \HOLTokenTransEnd \HOLTyOp{bool}}'',
i.e. functions taking CCS terms and returning sets of
variables (of type \HOLinline{\ensuremath{\alpha}}).
% Milner didn't give their definitions explicitly. However, one can
% imagine that, for most CCS operators they just pass into
% subterms without collecting anything (and $\nil$ gives $\emptyset$).
For their definition the interesting cases are those of 
recursion and agent variables, shown below  (here \texttt{DELETE}
and \texttt{INSERT} are set-theoretic operators of HOL's
\texttt{pred_set} theory):
\begin{center}
\begin{tabular}{|l|l|}
\hline
\HOLConst{FV} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLTokenDefEquality{} \HOLTokenLeftbrace{}\HOLFreeVar{X}\HOLTokenRightbrace{} & \HOLConst{FV} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{p}\ensuremath{)} \HOLTokenDefEquality{} \HOLConst{FV} \HOLFreeVar{p} \HOLConst{DELETE} \HOLFreeVar{X} \\
\HOLConst{BV} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLTokenDefEquality{} \HOLSymConst{\HOLTokenEmpty{}} & \HOLConst{BV} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{p}\ensuremath{)} \HOLTokenDefEquality{} \HOLFreeVar{X} \HOLConst{INSERT} \HOLConst{BV} \HOLFreeVar{p} \\
\hline
\end{tabular}
\end{center}
Furthermore, $E$ is a process, written \HOLinline{\HOLConst{IS_PROC} \HOLFreeVar{E}}, if
it does not contain any free variable, i.e. $\fv{E} = \emptyset$:
\begin{alltt}
   \HOLConst{IS_PROC} \HOLFreeVar{E} \HOLTokenDefEquality{} \HOLConst{FV} \HOLFreeVar{E} \HOLSymConst{\ensuremath{=}} \HOLSymConst{\HOLTokenEmpty{}}\hfill{[IS_PROC_def]}
\end{alltt}
And a list of CCS processes can be asserted by \HOLinline{\HOLConst{ALL_PROC}} defined upon \HOLinline{\HOLConst{IS_PROC}}:
\begin{alltt}
   \HOLConst{ALL_PROC} \HOLFreeVar{Es} \HOLTokenDefEquality{} \HOLConst{EVERY} \HOLConst{IS_PROC} \HOLFreeVar{Es}\hfill{[ALL_PROC_def]}
\end{alltt}

The set of free and bound
variables of a term need not be disjoint, e.g.\ in  $X + \recu X E$.
More importantly, when going
from a CCS expression to its sub-expressions, the set of free
variables may increase, while the set of bound variables either
remains the same or decreases. For instance, $\fv{\recu X (\mu.X)} = \emptyset$, while
$\fv{\mu.X} = \{X\}$. This property of $\fv{\cdot}$ brings some
difficulities when doing proofs about CCS transitions. As an evidence,
we prove the following fundamental result about $\fv{\cdot}$\hl{\mbox{~\citep[p.~1209]{milner1990operational}}}:

\begin{proposition}
\label{prop:transFV}
The derivatives of a process are themselves processes, i.e.
if $E \arr{\mu} E'$ and $\fv{E} = \emptyset$, then $\fv{E'} =
\emptyset$. Formally:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{E} \HOLBoundVar{u} \ensuremath{\HOLBoundVar{E}\sp{\prime}}. \HOLBoundVar{E} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{IS_PROC} \HOLBoundVar{E} \HOLSymConst{\HOLTokenImp{}} \HOLConst{IS_PROC} \ensuremath{\HOLBoundVar{E}\sp{\prime}}\hfill{[TRANS_PROC]}
\end{alltt}
\end{proposition}

\begin{proof}
We \hl{actually prove} a stronger result,  \hl{asserting} that the set of free
variables \hl{in a CCS process} may decrease through its derivatives:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{E} \HOLBoundVar{u} \ensuremath{\HOLBoundVar{E}\sp{\prime}}. \HOLBoundVar{E} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{FV} \ensuremath{\HOLBoundVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{FV} \HOLBoundVar{E}\hfill{[TRANS_FV]}
\end{alltt}

In textbooks such as \cite{Mil89} the proof of the above property is 
commented as  ``an easy action induction''.
In HOL, the proof 
 is actually non-trivial:
 ``action induction''  becomes a form of  higher-order application
of the following \emph{induction principle}, generated together with the SOS
rules (and that essentially says that \HOLinline{\HOLConst{TRANS}} is the smallest relation
satisfying the SOS rules):

@@

is completed.
%This is the essence of action induction.
Now, the  goal can be reduced to several conjunct

@@

the following goal left (the term above the dash line is the goal,
those below the line are assumptions):

@@

Note that \HOLinline{\HOLConst{FV}\\\;\HOLFreeVar{E}\\\;\HOLConst{DELETE}\\\;\HOLFreeVar{X}\\\;\HOLSymConst{\ensuremath{=}}\\\;\HOLConst{FV}\\\;\ensuremath{(}\HOLConst{rec}\\\;\HOLFreeVar{X}\\\;\HOLFreeVar{E}\ensuremath{)}}, so

@@

would be complete. Thus it remains to show that

@@

Proposition~\ref{prop:transFV} \hl{is essential in the proofs of the
  multivariable versions of all unique-solution
  theorems.
In comparison with} the above proof, the proof of the analogous
result for bound variables is  easier:
\begin{proposition}
\label{prop:transBV}
if $E \arr{\mu} E'$ then $\bv{E'} \subseteq \bv{E}$, or formally:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{E} \HOLBoundVar{u} \ensuremath{\HOLBoundVar{E}\sp{\prime}}. \HOLBoundVar{E} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{BV} \ensuremath{\HOLBoundVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{BV} \HOLBoundVar{E}\hfill{[TRANS_BV]}
\end{alltt}
\end{proposition}
\begin{proof}
Following the proof of Prop.~\ref{prop:transFV}, we do action
induction on $E \arr{\mu} E'$ and simplify all subgoals by set-theoretic
facts. The only subgoal left is:
\begin{alltt}
        \HOLinline{\HOLConst{BV} \ensuremath{\HOLFreeVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLFreeVar{X} \HOLConst{INSERT} \HOLConst{BV} \HOLFreeVar{E}}
   ------------------------------------
    0.  \HOLinline{\HOLConst{BV} \ensuremath{\HOLFreeVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{BV} \ensuremath{(}\ensuremath{[}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)}}
\end{alltt}
Note the difference with the proof of Prop.~\ref{prop:transFV}: now
we have \texttt{INSERT} in place of \texttt{DELETE}. Then, we directly
use the assumption and the transitivity of $\subseteq$ and try to
prove the following goal instead:
\begin{alltt}
        \HOLinline{\HOLConst{BV} \ensuremath{(}\ensuremath{[}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)} \HOLSymConst{\HOLTokenSubset{}} \HOLFreeVar{X} \HOLConst{INSERT} \HOLConst{BV} \HOLFreeVar{E}}
   ------------------------------------
    0.  \HOLinline{\HOLConst{BV} \ensuremath{\HOLFreeVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{BV} \ensuremath{(}\ensuremath{[}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)}}
\end{alltt}
This is an easy goal, due to the following lemma which is
 proved by induction on $E$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{X} \HOLBoundVar{E} \ensuremath{\HOLBoundVar{E}\sp{\prime}}. \HOLConst{BV} \ensuremath{(}\ensuremath{[}\ensuremath{\HOLBoundVar{E}\sp{\prime}}\ensuremath{/}\HOLBoundVar{X}\ensuremath{]} \HOLBoundVar{E}\ensuremath{)} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{BV} \HOLBoundVar{E} \HOLSymConst{\HOLTokenUnion{}} \HOLConst{BV} \ensuremath{\HOLBoundVar{E}\sp{\prime}}\hfill{[BV_SUBSET]}
\end{alltt}
Taking $E ' = \recu X E$ in the above lemma, we have
    \HOLinline{\HOLConst{BV}\\\;\ensuremath{(}\ensuremath{[}\HOLConst{rec}\\\;\HOLFreeVar{X}\\\;\HOLFreeVar{E}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]}\\\;\HOLFreeVar{E}\ensuremath{)}\\\;\HOLSymConst{\HOLTokenSubset{}}\\\;\HOLConst{BV}\\\;\HOLFreeVar{E}\\\;\HOLSymConst{\HOLTokenUnion{}}\\\;\HOLConst{BV}\\\;\ensuremath{(}\HOLConst{rec}\\\;\HOLFreeVar{X}\\\;\HOLFreeVar{E}\ensuremath{)}},
    the right side equals to \HOLinline{\HOLConst{BV}\\\;\HOLFreeVar{E}\\\;\HOLSymConst{\HOLTokenUnion{}}\\\;\ensuremath{(}\HOLFreeVar{X}\\\;\HOLConst{INSERT}\\\;\HOLConst{BV}\\\;\HOLFreeVar{E}\ensuremath{)}} $=$
    \HOLinline{\HOLFreeVar{X}\\\;\HOLConst{INSERT}\\\;\HOLConst{BV}\\\;\HOLFreeVar{E}}. This
completes
the    proof.
\end{proof}

@@

There are two ways to represent  the \multivariate
substitution  $E\{\til P/\til X\}$, that replaces each \hl{free occurrence}
of  variables $X_i$ in $E$ with the corresponding
$P_i$: (1) iterately applying the existing \univariate
\HOLinline{\HOLConst{CCS_Subst}}; (2) defining a new \multivariate substitution
function.
% In any case, we (reasonably) assume that all $X_i$ are
%distinct with each other. 
Note that  there is the
possibility that $\til P$   syntactically  again contains variables from $\til X$, thus
different orders of iterated substitutions may lead to different results.
% In fact, there are all kinds of issues with
% substitutions and applying them to term-like structures. 
To obtain the
maximal flexibility with a reduced effort, we defined
a \multivariate version of \HOLinline{\HOLConst{CCS_Subst}}
called \HOLinline{\HOLConst{CCS_SUBST}} based HOL's \texttt{finite\_map} theory~\cite{holdesc}.

@@

The function \HOLinline{\HOLConst{CCS_SUBST}} takes a finite
map \HOLinline{\HOLFreeVar{fm}} of type \HOLinline{\ensuremath{\alpha} \HOLTokenMapto{} \ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS}} and a CCS expression,
and returns another CCS expression in which
all occurrences of variables in \HOLinline{\HOLFreeVar{fm}} domain are substituted with
the corresponding value in \HOLinline{\HOLFreeVar{fm}}. Such a finite map can be built from
the list of variables \HOLinline{\HOLFreeVar{Xs}} and the corresponding targets
\HOLinline{\HOLFreeVar{Ps}}  by a helper function \HOLinline{\HOLConst{fromList}} (whose details 
are omitted here), while
\HOLinline{\HOLConst{CCS_SUBST}\\\;\ensuremath{(}\HOLConst{fromList}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Ps}\ensuremath{)}\\\;\HOLFreeVar{E}} is
abbreviated as \HOLinline{\ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]}\\\;\HOLFreeVar{E}}.
The substitution mechanism is
 order-independent. For most CCS operators it
just recursively calls itself on subterms. The only
interesting cases are at agent variables and recursion:
\begin{alltt}
   \HOLConst{CCS_SUBST} \HOLFreeVar{fm} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLTokenDefEquality{} \HOLKeyword{if} \HOLFreeVar{X} \HOLSymConst{\HOLTokenIn{}} \HOLConst{FDOM} \HOLFreeVar{fm} \HOLKeyword{then} \HOLFreeVar{fm} \HOLConst{'} \HOLFreeVar{X} \HOLKeyword{else} \HOLConst{var} \HOLFreeVar{X}\hfill{[CCS_SUBST_var]}
